\Title

### 2022-11-30 Grimaldi-etal-BiolCybernetics.tex
2.2

This raster plot is generated by spiking neurons which are defined on the one hand by the equations governing the evolution of its membrane potential dynamics on its soma and on the other hand by the characterization of the synaptic contacts on its dendritic tree, or on the sensory processes (such as the pixel on the sensor of a camera, or the photoreceptors in the retina). A classical characterization consists in detailing the synaptic weights of each synaptic contact on the dendritic tree, the so-called weight matrix. A more detailed description of these synaptic contacts reveals the added importance of heterogeneous delays, i.e., the precise timing from one afferent neuron's firing to its arrival in the soma, and how this changes the network's dynamics~\citep{izhikevich_polychronization_2006}. In that description, we can parameterize each neuron by the set of tuples defining both the weight and the delay of each synaptic contact. As a consequence, a set of input presynaptic spikes $\event$ will be processed by the dendrites defined by this synaptic set and notably by the respective delays, which will multiplex in time all events.

Let's formalize a layer of spiking neurons with heterogeneous delays (HD-SNN). Each neuron $\postsynaddr \in \postsynaddrspace$  connects to presynaptic afferent from $\presynaddrspace$. In biology, a single cortical neuron has generally several thousands of synapses, and each synapse may be defined by its synaptic weight and its delay, that is, the time it takes for one spike to travel from the presynaptic neuron's soma to that of the postsynaptic neuron. A postsynaptic neuron $\postsynaddr \in \postsynaddrspace$ is then described by the synaptic weights connecting it to a presynaptic afferent from $\presynaddrspace$ but also by the set of possible delays. Note that a neuron may contact an afferent neuron with multiple different delays. For each neuron $\postsynaddr$, we thus define a set of $\Nsyn^\postsynaddr$ synapses, as  $\synapse^\postsynaddr = \{(\presynaddr^\postsynaddr_\ranksyn, \synapticweight^\postsynaddr_\ranksyn, \synapticdelay^\postsynaddr_\ranksyn)\}_{\ranksyn \in [1,\Nsyn^\postsynaddr]}$, where each synapse $\synapse^\postsynaddr_\ranksyn$ is associated to a weight $\synapticweight^\postsynaddr_\ranksyn$, a delay $\synapticdelay^\postsynaddr_\ranksyn$ and a presynaptic address $\presynaddr^\postsynaddr_\ranksyn$. This defines the full connectivity of the heterogeneous model. As a consequence, an event stream $\event_\presynaddrspace$ on the presynaptic address space is transformed by the synapse $\ranksyn$ of neuron $\postsynaddr$ into a new event stream $\event_{\presynaddrspace, \ranksyn}$ which is defined by 
$\{ (\presynaddr_\arank = 
 \presynaddr_\ranksyn^\postsynaddr, \timev=\timev_\arank + 
\synapticdelay_\ranksyn^\postsynaddr )_{ \arank \in [1,\numevent]\text{, }\ranksyn \in [1,\Nsyn^\postsynaddr]} )
\}$. This new stream of events \note{formaliser le multiplexage}
When post-synaptic neurons are activated on their soma by this spatio-temporal motif the discharge probability will increase, notably when these spikes converge on the soma in a synchronous manner.

Taking the argument the other way around, one may form a generative model for generic raster plots. Indeed, any spike in the presynaptic address space is generated by sensory neurons (for instance photoreceptors in the retina, sensors in a CMOS chip) or by afferent spiking neurons. In the latter case, these are connected to the spiking cell by a set of weights and delays, whose structure is stable relatively to the coding timescale. When these connections are high and sparsely distributed, this firing will cause a specific temporal motif. Another example is given for the barn owl auditory system: As it hears the sound of a mouse, this sound will generate a specific spiking response in both ears, and specifically, the precise timing between the signal generated by the left relative to the right ear can be for instance used to determine the position of the prey~\citep{goodman_spike-timing-based_2010}. Overall, these examples show that raster plots may be considered as a mixture of the effects of different elementary causes, and that each event triggers a specific spatio-temporal motif. 
%
\subsubsection{Detecting spiking motifs}
%: Detection model

@@

From the perspective of simulating such event-based computations on standard chips, it is useful to transform this sparse representation into a dense representation. As such, we may first write any event-based input as the boolean matrix $A \in \{0, 1 \}^\presynaddrspace$. In this simplified model, we will consider that heterogeneous delays are limited in range such that the synaptic set can be represented by the dense matrix $\kernel^\postsynaddr$ giving for each neuron $\postsynaddr$ the weights as a function of presynaptic address and delay: $\forall {\ranksyn \in [1,\Nsyn^\postsynaddr]}, \kernel^\postsynaddr(\presynaddr_\ranksyn^\postsynaddr, \synapticdelay_\ranksyn^\postsynaddr) = \synapticweight_\ranksyn^\postsynaddr$. 
The probability of firing of a neuron $a$ at a given time $t$ can be understood as a Bernoulli trial whose (only) parameter is a bias $p(t, a) \in [0, 1]$. Assuming that the presence of PGs conditions the probability on all efferents, the logit (inverse of the sigmoid) of this probability bias can be written as the sum of the logit of each of these factors, whose values are given by the corresponding weights. PGs may be activated independently at random times and  we write that $B(b, t)=1$ if $b$ is activated at $t$ (and else $B(b, t)=0$). We can thus write the probability bias as the accumulated evidence given these factors as 

@@

%\subsubsection{Implementation using spiking neurons with heterogeneous delays}
%
The activation function of our spiking neural is a softmax function implementing a form of  Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. 
%It transforms this list of weights into a probability with the following formula:
%$
%Pr(k=\postsynaddr \; \vert \; \timev) =
%\frac 1 Z
%{\exp  (\mathcal{C}^\postsynaddr(\timev) +\bias^\postsynaddr) }
%$ 
%where $\mathcal{C}^\postsynaddr(\timev) = \sum
%\activeweights^\postsynaddr(t)
%$ is the sum of the synaptic weights and $\bias^\postsynaddr$ is the bias linked to neuron $\postsynaddr$. 
%In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike.
%%
%
\note{say that it is effortless in biology but difficult on conventional computers}
In our MLR model with $\Nclass=\Nspeed$ classes, a probability value is predicted for each event at address $\presynaddr_\arank$ and at time $\timev_\arank$ as a softmax function of the linear combination of the list of events on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the heterogeneous delays model. 
%
\subsubsection{Performance on synthetic data}
%
\begin{figure}%[t!]
    \centering
    \includegraphics[width=0.490\linewidth]{figures/THC_N_PGs.pdf}
    \includegraphics[width=0.490\linewidth]{figures/THC_N_PG_time.pdf}
	    \caption{Detecting event-based motifs using spiking neurons with heterogeneous delays. 
	    {\bf (a)}~Accuracy of PG detection as a function of the number $M$ of kernels.
	    {\bf (b)}~Accuracy of PG detection as a function of the temporal depth $D$ of kernels among $M=144$ kernels.
	    	     }
    \label{fig:model_results}
\end{figure}
% 
This generative model defined above allows to determine this inference model for guessing sources $B$ when observing a raster plot $A$. This assumes that we know the PGs as defined by the $W_b$ matrices. The underlying metric is the binary cross-entropy, as used in the logistic regression model. In particular, if we consider PG kernels with similar decreasing exponential time profile, we can prove that this is similar to the method of~\citet{berens_fast_2012}. In our specific case, the difference is that the regression is performed in both dendritic and delay space by extending the summation using a temporal convolution operator. Using this forward model, it is possible to estimate the logit (inverse of a sigmoid) $\hat{B}(b, t)$ for the presence of a PG of address $b$ and at time $t$ by using the transpose convolution operator. It thus comes that when observing $A$, then one may infer $\hat{B} = A \ast W^T$ and select the most activated items.

To quantify the efficiency of this operation, we generated $M=55$ synthetic PGs as random independent kernels over $128$ presynaptic inputs and $D=127$ possible delays. We drew random independent instances of $B$ with a length of $T=1000$ time steps and with on average $2.0$ occurrences each. This allowed us to generate raster plots which we use to infer $\hat{B}$. We compute the accuracy as the rate of true positive detections (both for inferring the address and its exact timing) and observe on average $\approx 98\%$ correct detections. We further extended this result by showing how the accuracy would evolve as a function of the number of simultaneous PGs, while keeping the same frequency of occurrence. We show in Figure~\ref{fig:model_results}~(a) that the accuracy of finding the right PG is still above $80\%$ accuracy with more than $1364$ overlapping PGs. Moreover, we show in Figure~\ref{fig:model_results}~(b) that (with $M=55$ PGs fixed) the accuracy increases notably as the temporal depth $D$ of the PG kernel increased, demonstrating quantitatively the potential of heterogeneous delays. These results were obtained while assuming that we know $W$. However, this is in general not the case, for instance when observing the raster plot of a population of neurons. In the following, we will define a generic visual task and determine a learning algorithm to solve it. %Inspired by the k-means algorithm, it is possible to devise a self-supervised learning algorithm. Our preliminary results show that it is possible to retrieve PGs embedded in the data, yet that further analysis is necessary to improve the convergence of the algorithm. In particular, it seems promising to use a sparseness constraint in the inference mechanism such as to remove spurious correlations in the inference.

>>>>

% This raster plot is generated by spiking neurons which are defined on the one hand by the equations governing the evolution of its membrane potential dynamics on its soma and on the other hand by the characterization of the synaptic contacts on its dendritic tree, or on the sensory processes (such as the pixel on the sensor of a camera, or the photoreceptors in the retina). A classical characterization consists in detailing the synaptic weights of each synaptic contact on the dendritic tree, the so-called weight matrix. A more detailed description of these synaptic contacts reveals the added importance of heterogeneous delays, i.e., the precise timing from one afferent neuron's firing to its arrival in the soma, and how this changes the network's dynamics~\citep{izhikevich_polychronization_2006}. In that description, we can parameterize each neuron by the set of tuples defining both the weight and the delay of each synaptic contact. As a consequence, a set of input presynaptic spikes $\event$ will be processed by the dendrites defined by this synaptic set and notably by the respective delays, which will multiplex in time all events.

Let's formalize a layer of spiking neurons with heterogeneous delays (HD-SNN). Each neuron $\postsynaddr \in \postsynaddrspace$  connects to presynaptic afferent from $\presynaddrspace$. In biology, a single cortical neuron has generally several thousands of synapses. Each synapse may be defined by its synaptic weight and its delay, that is, the time it takes for one spike to travel from the presynaptic neuron's soma to that of the postsynaptic neuron. A postsynaptic neuron $\postsynaddr \in \postsynaddrspace$ is then described by the synaptic weights connecting it to a presynaptic afferent from $\presynaddrspace$ but also by the set of possible delays. Note that a neuron may contact an afferent neuron with multiple different delays. For each neuron $\postsynaddr$, we thus define a set of $\Nsyn^\postsynaddr$ synapses, as  $\synapse^\postsynaddr = \{(\presynaddr^\postsynaddr_\ranksyn, \synapticweight^\postsynaddr_\ranksyn, \synapticdelay^\postsynaddr_\ranksyn)\}_{\ranksyn \in [1,\Nsyn^\postsynaddr]}$, where each synapse $\synapse^\postsynaddr_\ranksyn$ is associated to a weight $\synapticweight^\postsynaddr_\ranksyn$, a delay $\synapticdelay^\postsynaddr_\ranksyn$ and a presynaptic address $\presynaddr^\postsynaddr_\ranksyn$. This defines the full connectivity of the HD-SNN model. As a consequence, an event stream $\event_\presynaddrspace$ on the presynaptic address space is transformed by the synapse $\ranksyn$ of neuron $\postsynaddr$ into a new event stream $\event_{\presynaddrspace, \ranksyn}$ which is defined by 
$\{ (\presynaddr_\arank = 
 \presynaddr_\ranksyn^\postsynaddr, \timev=\timev_\arank + 
\synapticdelay_\ranksyn^\postsynaddr )_{ \arank \in [1,\numevent]\text{, }\ranksyn \in [1,\Nsyn^\postsynaddr]} )
\}$. This new stream of events \note{formaliser le multiplexage}
When post-synaptic neurons are activated on their soma by this spatio-temporal motif the discharge probability will increase, notably when these spikes converge on the soma in a synchronous manner.

@@

\note{ éclaircir }

Taking the argument the other way around, one may form a generative model for generic raster plots. Indeed, any spike in the presynaptic address space is generated by sensory neurons (for instance photoreceptors in the retina, sensors in a CMOS chip) or by afferent spiking neurons. In the latter case, these are connected to the spiking cell by a set of weights and delays, whose structure is stable relatively to the coding timescale. When these connections are high and sparsely distributed, this firing will cause a specific temporal motif. Another example is given for the barn owl auditory system: As it hears the sound of a mouse, this sound will generate a specific spiking response in both ears, and specifically, the precise timing between the signal generated by the left relative to the right ear can be for instance used to determine the position of the prey~\citep{goodman_spike-timing-based_2010}. Overall, these examples show that raster plots may be considered as a mixture of the effects of different elementary causes, and that each event triggers a specific spatio-temporal spiking motif. 
%
\subsubsection{Detecting spiking motifs}
%: Detection model

From the perspective of simulating such event-based computations on standard CPU- or GPU-based computers, it is useful to transform this event-based representation into a dense representation. Indeed, we may transform any event-based input as the boolean matrix $A \in \{0, 1 \}^{N\times T}$, where $N$ is the number of neurons in $\presynaddrspace$ and $T$ is the number of time bins. In this simplified model, we will consider that heterogeneous delays are limited in range such that the synaptic set can be represented by the dense matrix $\kernel^\postsynaddr$ giving for each neuron $\postsynaddr$ the weights as a function of presynaptic address and delay: $\forall {\ranksyn \in [1,\Nsyn^\postsynaddr]}, \kernel^\postsynaddr(\presynaddr_\ranksyn^\postsynaddr, \synapticdelay_\ranksyn^\postsynaddr) = \synapticweight_\ranksyn^\postsynaddr$. 
The probability of firing of a neuron $a$ at a given time $t$ can be understood as a Bernoulli trial whose (only) parameter is a bias $p(t, a) \in [0, 1]$. Assuming that the presence of spiking motifs conditions the probability on all efferents, the logit (inverse of the sigmoid) of this probability bias can be written as the sum of the logit of each of these factors, whose values are defined by the corresponding weights. Spiking motifs may be activated independently and at random times, such that  we write this activity as $B(b, t)=1$ if $b$ is activated at $t$ and else $B(b, t)=0$. We can thus write the probability bias as the accumulated evidence given these factors as 

@@

% 
This generative model defined above allows to determine this inference model for guessing sources $B$ when observing a raster plot $A$. This assumes that we know the spiking motifs as defined by the $W_b$ matrices. The underlying metric is the binary cross-entropy, as used in the logistic regression model. In particular, if we consider kernels with similar decreasing exponential time profile, we can prove that this is similar to the method of~\citet{berens_fast_2012}. In our specific case, the difference is that the regression is performed in both dendritic and delay space by extending the summation using a temporal convolution operator. Using this forward model, it is possible to estimate the logit (inverse of a sigmoid) $\hat{B}(b, t)$ for the presence of a PG of address $b$ and at time $t$ by using the transpose convolution operator. It thus comes that when observing $A$, then one may infer $\hat{B} = A \ast W^T$ and select the most activated items.

%\subsubsection{Implementation using spiking neurons with heterogeneous delays}
%
% The activation function of our spiking neural is a softmax function implementing a form of  Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. 
%It transforms this list of weights into a probability with the following formula:
%$
%Pr(k=\postsynaddr \; \vert \; \timev) =
%\frac 1 Z
%{\exp  (\mathcal{C}^\postsynaddr(\timev) +\bias^\postsynaddr) }
%$ 
%where $\mathcal{C}^\postsynaddr(\timev) = \sum
%\activeweights^\postsynaddr(t)
%$ is the sum of the synaptic weights and $\bias^\postsynaddr$ is the bias linked to neuron $\postsynaddr$. 
%In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike.
%%
%
% \note{say that it is effortless in biology but difficult on conventional computers}
% In our MLR model with $\Nclass=\Nspeed$ classes, a probability value is predicted for each event at address $\presynaddr_\arank$ and at time $\timev_\arank$ as a softmax function of the linear combination of the list of events on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the heterogeneous delays model. 
%
\subsubsection{Performance on synthetic data}
%
\begin{figure}%[t!]
    \centering
    \includegraphics[width=0.490\linewidth]{figures/THC_N_PGs.pdf}
    \includegraphics[width=0.490\linewidth]{figures/THC_N_PG_time.pdf}
	    \caption{Detecting event-based motifs using spiking neurons with heterogeneous delays. 
	    {\bf (a)}~Accuracy of detection as a function of the number of kernels.
	    {\bf (b)}~Accuracy of detection as a function of the temporal depth $D$ of kernels among $M=144$ kernels.
	    }
    \label{fig:model_results}
\end{figure}

To quantify the efficiency of this operation, we generated $M=144$ synthetic spiking motifs as random independent kernels over $128$ presynaptic inputs and $D=71$ possible delays. We drew random independent instances of $B$ with a length of $T=1000$ time steps and with on average $1.0$ occurrences in each draw. This allowed us to generate raster plots which we use to infer $\hat{B}$. We compute the accuracy as the rate of true positive detections (both for inferring the address and its exact timing) and observe on average $\approx 98\%$ correct detections. We further extended this result by showing how the accuracy would evolve as a function of the number of simultaneous PGs, while keeping the same frequency of occurrence. We show in Figure~\ref{fig:model_results}~(a) that the accuracy of finding the right PG is still above $80\%$ accuracy with more than $1364$ overlapping PGs. Moreover, we show in Figure~\ref{fig:model_results}~(b) that (with $M=144$ PGs fixed) the accuracy increases notably as the temporal depth $D$ of the PG kernel increased, demonstrating quantitatively the computational advantage of using heterogeneous delays. These results were obtained while assuming that we know $W$. However, this is in general not the case, for instance when observing the raster plot of a population of neurons. In the following, we will define a generic visual task and determine a learning algorithm to solve that challenging task. %Inspired by the k-means algorithm, it is possible to devise a self-supervised learning algorithm. Our preliminary results show that it is possible to retrieve PGs embedded in the data, yet that further analysis is necessary to improve the convergence of the algorithm. In particular, it seems promising to use a sparseness constraint in the inference mechanism such as to remove spurious correlations in the inference.
